Scaling Automated Theory Exploration

### 2015-08-06 AI4FM/article.tex
Improvements

More generally, automated TE is an \emph{optimisation problem}; an area which
has been studied extensively in the fields of Artificial Intelligence and
Machine Learning. For a more thorough treatment, \cite{geng2006interestingness}
surveys recent work on search algorithms and \cite{blum2011hybrid} surveys
interestingness criteria in the context of data mining.

@@

Automated theory exploration has been applied to libraries in Isabelle
and Haskell, although we focus on the latter as its implementations are
the most mature (demonstrated by \textsc{Hipster}, which explores Isabelle by
translating code to Haskell first). Haskell is interesting to target,
since its emphasis on pure functions and algebraic datatypes make equational
properties common; however, Haskell is not a theorem prover, and hence it cannot
represent such as universally quantified equations. Also, the widespread use of
recursion and higher-order functions makes automated deduction non-trivial
effort is spent discovering and proving these properties compared to
proof-oriented systems like Isabelle.

Due to Haskell's relative popularity, there are large code repositories
such as \textsc{Hackage} available to explore, with the potential to
benefit existing library authors and users in comprehending and
maintaining their code \cite{QuickSpec}.

@@

\begin{itemize}
\item
  Expression enumeration is brute-force; this could scale to larger
  terms and theories using a heuristic algorithm.
\item
  ``Interestingness'' is a fixed part of the algorithm: an equation is
  interesting if it cannot be derived from previous equations. As we
  increase the size of our theory, this becomes unsatisfying in two
  ways:

  \begin{itemize}
  \item
    The number of irreducible equations grows, making it desirable to
    impose extra conditions of a more subjective nature.
  \item
    Surprising, insightful equations may be discarded, if they are
    actually reducible in some complex, non-obvious way. A more
    subjective interestingness measure could be used to veto such
    rejections.
  \end{itemize}
\item
  The system does not propose candidate equations by data mining
  previous results; generalisation methods like anti-unification could
  do this, and at the same time remove the requirement that general
  forms must be enumerated early.
\item
  All type-safe combinations of the given expressions are tried, whilst
  it may be discernable a priori that some combinations are not worth
  considering (either because they are never related, or because their
  relations are never interesting). A pre-processor could make large
  theories more tractable by selecting combinations which are likely to
  be related, similar to premise selection in automated theorem proving
  \cite{kuhlwein2012overview}.
\end{itemize}

@@

We are implementing a system called \textsc{ML4HS} to investigate these
ideas. Our initial hypothesis that expressions with similar definitions
are more likely to be related by equational properties than those
without, and hence a similarity-based clustering method such as that of
\textsc{ML4PG} \cite{journals/corr/abs-1302-6421} can be used to implement
a divide and conquor pre-processor.

Since our aim is to scale up theory exploration, we treat entire
\textsc{Hackage} packages as our theories. \textsc{ML4HS} will manage
downloading, compiling, managing dependencies, etc. automatically. This
both eliminates the need to define theories manually, and may be useful
in its own right as a mechanism to execute arbitrary Haskell code from
arbitrary modules in arbitrary packages.

>>>>

@@

Automated theory exploration has been applied to libraries in Isabelle
and Haskell, although we focus on the latter as its implementations are
the most mature (demonstrated by the fact that \textsc{Hipster} explores
Isabelle by first translating it to Haskell). Haskell is interesting to target,
since its use of pure functions and algebraic datatypes causes many functions to
follow algebraic laws; however, since Haskell's type system cannot encode such
laws (without difficulty \cite{lindley2014hasochism}, at least), less
effort is given to finding and stating them; compared to a theorem proving
setting like Isabelle, where these laws are often the main focus.

Due to Haskell's relative popularity, there are also large code repositories
such as \textsc{Hackage} available to explore. This gives several potential
benefits: in the short term, existing library authors and users may be assisted
in comprehending and maintaining their code \cite{QuickSpec}; more notably, TE
could be used to data mine new abstractions, such as typeclasses, with relevance
to real-world code.

@@

For a more thorough treatment,
surveys recent work on search algorithms and  surveys
interestingness criteria in the context of data mining.

\begin{itemize}
\item
  Enumerating expressions is a brute-force approach. Identifying more scalable
  alternatives to brute-force search is a well-studied area of Artificial
  Intelligence and Machine Learning. In particular, there is potential in
  applying heuristic algorithms like those surveyed in \cite{blum2011hybrid}.
\item
  Many ``interestingness'' criteria are surveyed in
  \cite{geng2006interestingness}, but existing systems are tightly coupled to
  one particular notion. Whilst pre- and post-processors could provide
  additional filtering, there is merit in building a more loosely-coupled
  alternative, for two reasons:

  \begin{itemize}
  \item
    As the number of irreducible equations grows, it may be desirable to
    impose extra conditions of a more subjective nature.
  \item
    Surprising, insightful equations may be discarded by current systems, if
    they are actually reducible in some complex, non-obvious way. A more
    subjective interestingness measure could be used to veto such
    rejections.
  \end{itemize}
\item
  HipSpec does not propose candidate equations by data mining
  previous results; generalisation methods like anti-unification could
  do this, and at the same time remove the requirement that general
  forms must be enumerated early.
\item
  All type-safe combinations of the given expressions are tried, whilst
  it may be discernable a priori that some combinations are not worth
  considering (either because they are never related, or because their
  relations are never interesting).
\end{itemize}

We consider all of these worth investigating, although in this paper we
focus on the last item.

@@

We are implementing a system called \textsc{ML4HS} to investigate scalable
automated theory exploration for Haskell. At this stage, we provide a
pre-processor for \textsc{QuickSpec} which attempts to make exploration of large
theories more tractable by splitting them into a collection of smaller theories.

Inspired by premise selection in automated theorem proving
\cite{kuhlwein2012overview}, and similarity-based clustering method such as that
of \textsc{ML4PG} \cite{journals/corr/abs-1302-6421}, we are investigating the
hypothesis that similar expressions are more likely to be related by equational
properties than dissimilar expressions. This has lead to a divide and conquor
approach, shown in figure 1.

Since our aim is to scale up theory exploration, our pre-processor also
increases the level of automation compared to existing systems. We consider
\textsc{Hackage} packages to be theories, and treat package management,
downloading, compiling, dependency resolution, etc. as necessary tasks for a TE
system such as \textsc{ML4HS} to implement.

Not only does this eliminate the need to define theories manually, as a sub-set
of available expressions, but may also be useful in its own right as a mechanism
to execute arbitrary Haskell code from arbitrary modules in arbitrary packages.
